Discussion Draft

### 2020-04-14 doc/Discussion.rmd
more work on chromatin str

**if none of the models turn out to be appropriate,** it 
could be due to assumptions we made regaurding....

**We also note that** The musculus and molossinus subspecies have more shared ancestry (compared to domesticus) (cite Geraldes?), so we  cant determine if the rapid male evolution in gwrr are independant (mulitple events) or due to incomplete lineage sorting (single event).

If these hypotheses turn out to not be true, (a prediction that can be made, due to the rapid male specific

The consequences of this reversed heterochismy in house mice is that the sex-average / species average of gwRR is higher. meaning/resulting lower ld overall.

@@

If these hypotheses turn out to not be true, (a prediction that can be made, due to

The consequences of this difference in meitic events -- (it fits follows from some basic differences in reproductive stategies between males and females) -- females diversify offspring, males --

-- predictions for adaptations to changing environments -- but applied to short timescales -- since all the genes go back in the other sex the fllowing generation.

@@

If these hypotheses concerning evolution of these patterns are not true -- there's still predictions that can be made given the consequences of these patterns. The consequences of this difference in the typical landscape as reviewed in [@sardell_sex_2020]. For the sex averaged landscape along chromosomes, any diviation from uniform placement is generally due to male specific patterns.

@@

@@

longer female axis has evol

 consequences of longer female axis - 
 (normalzed interference strength due to axis differences / Chromtain loop length)

We also hypothesize that the sex differences in chromatin structure are closely connected to sex differences  in the interference strength and our results mirror those in [@petkov2007] and follow the prediction that shorter-loop will have stronger interference.

According to the tethered loop model of DSB formation (cite schmenti?), the 'real estate' for COs are the chromatin / DNA loops, the density of whch is conserved in the axis and loop formation (later the SC str), so the number of loops is equvilent to the length of the axis. **The chromosomes are converted to a 2D space (the axis length of a given chromosome is the limited factor for total CO number)**.  Only one of the models COM aligns with our observed results for sexual dimorphism for axis length. **COM** predicts that axis length will be larger in oocytes due to the larger volume.

@@

integrate physical models -- centromere and kinetochore

>

>>>>

0. Regardless of if these hypotheses turn out to be true, our results extend previous reports (dumont) of rapid male-specific evolution in multiple subspecies and strains of house mouse.

1. whether these instances of evolution are due to seperat independent events or segregating genetic variation within the population requires further study.

2. These results may hold importance to animal breeding programs -- may suggest that even for species with female biased heterochiasmy, selecting for higher male specific genome wide recombination rates may be the most efficient way to increase the sex-averaged recombination rates. This could be important application for animal breeding programs (Battagin et al 2016).

@@

0. Reguardless of whether these models and hypothese are true ... if the pattern of females having more across cell variation in gwRR

1. It has a consequence of fitting with distinct evolutionary streategies between the sexes of females diversifying the genetics their offspring and males limiting the shuffling of successful combinations (Trivers).

@@

If these hypotheses concerning evolution of these patterns **are not true** -- there's still predictions that can be made given the consequences of these patterns. The consequences of this difference in the typical landscape as reviewed in [@sardell_sex_2020]. For the sex averaged landscape along chromosomes, any diviation from uniform placement is generally due to male specific patterns.

@@

# significance, consequences
# weak evidence of sex differences in chromatin organization, mice-humans have a similar direction -- but At opposite -- not much empiracal evidence. (more data needed)

# a consequence of these sex differences for the chromatin organization is -- has an outcome of different effects on interference strenth measured in genetic map space -- for a given chromosome, chromatin organization of longer axis will result in shorter loops (emenating off the central axis)

#4.ppl should care -- this is an important aspect of -- how sexual dimorphism in recombination /linkage maps has been charaecterized -- but due to a lack of cytological evidence in both sexes /linking this to chromatin organization has only been done for a handful of species (mice, human, At? Dourad)

@@

0. There is weak evidence that female chromatin organization having a longer axis is a conserved trait. (only mice and humans, but there is a notable lack of data)

1. We note that A consequence of sexual dmorphism is the chromatin axis (chromatin organization) is that interference strength^norm^ / (general patterns of sex specific genetic maps) will be sexually dimorphic.

2. The density of tethering of loops to the axis is highly conserved (Zand Kleck) - 
variablibilty of the axis length of a given chromosome stems from varing the loop sizes (Z n K)).

3 When interference strength is translated into base pairs (Mb), the chromatin organization with larger loops will have stronger interference (Mb) since more base pairs are sequestered in the DNA loops per unit of SC/axis.

5. while this prediction / the general pattern -- relationhsip of chromatin organization and interference strength is complicated -- the biology of chromatin (3D chromatin organization, heterogeneity of loop size within (cite) and among chromosomes (cite) and (limitations in most / for identifying indiviual chromosomes across single cells)

(this prediction was outlined in Petkov, -- Lynn and Gruhn)

It holds for several model organzism (mice, human, At) and sexual dimorphism in linkage maps (has been one of the oldest measurements of heterochiasmy --  This is an important aspect of how sexual dimorphism in recombination linkage maps has been charaecterized but a lack of cytological evidence for both sexes --

--(so get the cytological data to test if this prediction holds)

<We also hypothesize that the sex differences in chromatin structure are closely connected to sex differences  in the interference strength and our results mirror those in [@petkov2007] and follow the prediction that shorter-loop will have stronger interference.  According to the tethered loop model of DSB formation (cite schmenti?), the 'real estate' for COs are the chromatin / DNA loops, the density of whch is conserved in the axis and loop formation (later the SC str), so the number of loops is equvilent to the length of the axis. **The chromosomes are converted to a 2D space (the axis length of a given chromosome is the limited factor for total CO number)**.  Only one of the models COM aligns with our observed results for sexual dimorphism for axis length. **COM** predicts that axis length will be larger in oocytes due to the larger volume.>

@@

integrate physical models -- centromere and kinetochore
>

Diversity of systems for overcoming limitations

1. identifying indiviual (autosomes) Chromosomes across cells: FISH (diffecult) or karyotypes with identifyable chromosomes (either into groups or largest)(birds, zfish?, peromyscus, Rbt mice, bovids?, humans, ), translocations (robertsonian translocations),

2. non-model orgs can be used to study meiosis -- look at the weird ones (protezoa)
